---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
Reproducible Research assignment 1: Daily activity monitoring
==============================================================

Note: Rstudio is being funny. The html was knit from this .Rmd file using the following:

```
library(knitr)
library(markdown)
knit('PA1_template.Rmd')
markdownToHTML('PA1_template.md','PA1_template.html')
```

## Loading and preprocessing the data
load the data here. Load the data from the csv as "data", casting the "date" field as a date class. Do some aggregation, calculate total steps per day and mean steps per interval. 
```{r}
data<-read.csv('activity.csv')
data$date <- as.Date(data$date)
data$weekday <- weekdays(data$date)

stepsPerDay<-aggregate(steps~date,data,sum)
stepsPerInterval<-aggregate(steps~interval,data,mean)

```
plot steps taken per day
```{r}
library(ggplot2)
g<-ggplot(data=stepsPerDay,aes(steps,fill=..count..),na.rm=TRUE) 
labels<- labs(x='steps per day')
g+  geom_histogram(binwidth=1000) + labels
```


## What is mean total number of steps taken per day?

```{r}
themean=mean(stepsPerDay$steps)
themedian=median(stepsPerDay$steps)
```
```{r}
themean
```
```{r}
themedian
```
That's pretty good. My phone says I did 65k steps last week.

## What is the average daily activity pattern?
Want a plot of steps per 5-min interval, averaged over all days
Use the ```aggregate``` function to get this data
```{r}
g<-ggplot(data=stepsPerInterval,aes(interval,steps,colour='red')) 
labels<- labs(x='interval',y='mean steps per 5 minute interval')
g+  geom_line() + labels
```

Which 5 minute interval, on average, contains the most steps?
```{r}
stepsPerInterval[which.max(stepsPerInterval$steps),]
```
interval 835 (206 steps)


## Imputing missing values

How many missing values are there?
```{r}
dim(data)
sum(is.na(data$steps))
```
Quite a bit. How are the missing values distributed by day? Group steps by day, sum the NA values for each group.
```{r}
suppressMessages(library(dplyr))
missByDay<- data %>% select(date,steps) %>% group_by(date) %>% summarise_each(funs(themiss=sum(is.na(.))))
missByDay %>% filter(themiss > 0)

```
So for each of these 8 days, the steps data for every interval is missing. That makes life easier, as we're not sampling some intervals more often than others.

Impute things. 

```{r}
stepsPerWeekday<-aggregate(steps~weekday,data,sum)
```
Compute the mean (median?) steps  grouped by weekday  and interval 
```{r}
data$weekday = weekdays(data$date)
lookup<- data %>% select(steps,weekday,interval) %>% group_by(weekday,interval) %>%
 summarize_each(funs(meansteps=mean(.,na.rm=TRUE),medsteps=median(.,na.rm=TRUE)))
```

plot the mean steps per weekday/interval

```{r}
g<- ggplot(aes(x=interval,y=meansteps,colour=weekday),data=lookup) + geom_line()
g
```

plot the median

```{r}
g<- ggplot(aes(x=interval,y=medsteps,colour=weekday),data=lookup) + geom_line()
g
```

Choose mean, just because.
Copy data to imputed dataframe

This question on [stackoverflow](http://stackoverflow.com/questions/35670213/replace-values-in-some-rows-based-on-other-dataframe-mapping-with-r) was helpful.

left join impData with lookup (keeps all rows of impData, pulls things from lookup)
copy meansteps to steps for rows where steps is NA, then drop the meansteps column

```{r}
thecolumn<-"meansteps"
impData<-data
impData <- impData  %>% left_join( select(lookup,c(interval,weekday,meansteps,medsteps)),by=c("weekday"="weekday","interval"="interval"))
impData[is.na(impData$steps),"steps"]<-impData[is.na(impData$steps),thecolumn]
impData <- impData %>% select(date,interval,weekday,steps)
```

make a histogram of total steps, same as for part 1 above, but using the imputed data
```{r}
stepsPerDayi<-aggregate(steps~date,impData,sum)
g<-ggplot(data=stepsPerDayi,aes(steps,fill=..count..),na.rm=TRUE) 
labels<- labs(x='steps per day',title='imputed data')
g+  geom_histogram(binwidth=1000) + labels
```

mean and median daily steps after imputation
```{r}
mean(stepsPerDayi$steps)
median(stepsPerDayi$steps)
```
These are a little different, because reasons



## Are there differences in activity patterns between weekdays and weekends?

scatterplot steps vs interval for weekdays and weekends

```{r}
impData$weekend<- factor(impData$weekday %in% c('Saturday','Sunday'),labels=c('Weekday','Weekend'))
g<- ggplot(data=impData,aes(x=interval,y=steps,colour=steps))
g+ geom_point() + facet_grid(.~weekend) + scale_colour_gradientn(colours=c('purple','red')) + geom_smooth()

```

Yup, there are differences. Subject tends to be more active earlier in the day during the week, and more active later in the day during the weekend.


